---
title: "vignette PlasmodeSim"
output: html_vignette
date: "2022-10-19"
---
Welcome to the vignette about the R package PlasmodeSim. This package is still under development.

```{r,begin , include= F}

# use default LASSO logistic regression

```

## installing plasmodeSim using remotes
To install using `remotes` run:
```{r, install, eval=F}
install.packages("remotes")
remotes::install_github("GidiusVanDeKamp/PlasmodeSim")
```


## Setting up
This documents skips some parts, we have skipped the steps to obtain the plpResults and the plpData. 

We begin by loading the package, together with the package `PatientLevelPrediction`, since the package `PlasmodeSim` is designed to be used with `PatientLevelPrediction`.
````{r, setup, message= F}
library(dplyr) 
library(PlasmodeSim)
library(PatientLevelPrediction)

modelSettings <- PatientLevelPrediction::setLassoLogisticRegression()

plpResult <- PatientLevelPrediction::loadPlpResult("~/R/internshipErasmusMC/simulate-new-patients-outcomes/plp_demolog/plpResult")
prediction <- plpResult$prediction %>% dplyr::filter(evaluationType == "Test")

plpData<- PatientLevelPrediction::loadPlpData("~/R/internshipErasmusMC/simulate-new-patients-outcomes/plp_demo/plpData" )
````
In this file we will show which functions are in the `PatientLevelPrediction` package, by adding `PatientLevelPrediction::` before the function. 

## Example 1

In this example we obtain new outcomes of a fitted logistic model.
```{r, ex1, message=F}
plpModel <- plpResult$model
newprobs <- PatientLevelPrediction::predictPlp(plpModel, plpData, plpData$cohorts)
```` 
The function predictPlp returned this information. 

`````{r}
newOut <- newOutcomes(200, newprobs)
head(newOut)

````
In the output of newOut patients are drawed randomly with the same chance, the patients could be drawed multiple times. If this happens they can have a different outcome. The function `newOutcomes` needs a data set where the column that contains the probabilities is called `value`.

## Example 2 
We here we show how to simulate new outcomes from an unfitted logistic model.

````{R, ex2,message=F } 
Parameters <- plpResult$model$model$coefficients
UnfittedParameters <- Parameters
UnfittedParameters[1,1] <- -0.4
UnfittedParameters[2:4,1] <- 0.4
head(UnfittedParameters)
````
For the logistic model it is necessary that the parameters are stored in a dataset with a column called `betas` and a column called `covariateIds`.  

````{r,message=F}
plpModelunfitted <- makeModel(UnfittedParameters, "logistic")
newprobs <- PatientLevelPrediction::predictPlp(plpModelunfitted, plpData,plpData$cohorts)
newOut <- newOutcomes(200, newprobs)
head(newOut)
````



## Visual simulations
The function `visualOutcome` simulated new data and then plots the frequency of the outcome.
Right now the function `visualOutcome` only works for a logistic model. The green line in the plots is the avegate outcome in the original dataset. 
````{r, visual, message= F, out.width = "60%"}
visualOutcome(plpData,50,200,Parameters)
visualOutcome(plpData,50,200,UnfittedParameters)
````
Here we have plotted 50 times the frequency of the outcome for a simulated dataset with 200 people. 

## Visual of a specific covariate

`````{r, visual2, message= F, out.width = "60%"}

covariateIdToStudy<- plpResult$covariateSummary$covariateId[3]
UnfittedParameters[3,]

visualOutcomeCovariateId(plpData, covariateIdToStudy, 20, 200, UnfittedParameters, "logistic")
visualOutcomeCovariateId2(plpData, covariateIdToStudy, 20, 200, UnfittedParameters, "logistic")

````

As one can see `visualOutcomeCovariateId` and `visualOutcomeCovariateId2` are very similiar, they both calculate and plot the frequency for a group with a specific covariate present. The small difference is that `visualOutcomeCovariateId` filters a newly simulated dataset set to only keep the patients where the covariate is present, and `visualOutcomeCovariateId2` only simulates new outcomes for patients that have the covariate present. 
We see they are almost the identical only `visualOutcomeCovariateId2` is spread out less because the groups for calculating the frequency with are bigger.

## survival times outcomes. 

We have to think about how to design this.
For simulating the survival times we need more than one probability, we use something like the **CDF**. 
When using `newPropsCoxversion2(plpModel, plpData)` we obtain a matrix of size *population size* by *number of unique times of the realizations* and a vector of length  *number of unique times of the realizations*. In the matrix we store the probabilities of surviving upto the times stored in the returned vector. 

````{r, cox model setup, results= 'hide'}
modelSettings <- PatientLevelPrediction::setCoxModel()
plpResult <- PatientLevelPrediction::loadPlpResult("~/R/internshipErasmusMC/simulate-new-patients-outcomes/plp_demoCox/plpResult")
plpModel<- plpResult$model
````

````{R, cox model}
plpModel$trainDetails$modelName

listpropandtime <- newPropsParametersPlpModel(plpModel, plpData)
newprops <- listpropandtime[[1]]
posibletimes <- listpropandtime[[2]]

NewOut <- newOutcomesSurvivalTimes(newprops, 30, posibletimes)
head(NewOut) # add the rowId's 
````
To simulate the outcomes we use the function `newOutcomesSurvivalTimes`. 
Right now this simulation has not added the censored data. also it does not uses the package survival. 

